---
title: "Optimal Visit Placement in Longitudinal Trials"
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{setspace}
  - \setstretch{1.1}
---

```{r setup, include=FALSE}
library(nlme)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

set.seed(20260305)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5.5,
  fig.height = 3.5,
  fig.align = "center",
  dev = "pdf"
)

options(scipen = 1, digits = 3)
```

# Introduction

In longitudinal clinical trials, subjects are measured repeatedly
over a fixed study period. The conventional approach assigns
observation times at equally spaced intervals (for example,
monthly or quarterly visits) largely for administrative
convenience rather than statistical efficiency. A growing body
of work in optimal experimental design demonstrates that the
placement of these observation times can substantially affect
the precision with which treatment effects are estimated.

## Optimal allocation of observation times

@winkensOptimalTimepointsClinical2005 provided the foundational
analysis of this problem for two-group trials with linearly
divergent treatment effects. They showed that the optimal
allocation of time-points depends critically on the assumed
covariance structure. Under compound symmetry (CS), efficiency
improves by clustering repeated measures near the end of the
study period, where the treatment effect is largest. Under a
first-order autoregressive (AR(1)) structure, only two
measurements -- at baseline and at study end -- are needed for
highly efficient estimation. Under a more realistic AR(1) plus
measurement error structure, the optimal spacing depends on the
specific covariance parameter values, but the equally spaced
design is generally suboptimal.

In a companion paper, @winkensOptimalNumberRepeated2006 extended
these results to address the joint optimization of the number
of repeated measures and group sizes under budget constraints.
They found that when budget is fixed and the covariance structure
includes an autoregressive component, two repeated measures per
subject often yield highly efficient treatment effect estimators.
Otherwise, intermediate measures can reduce the variance of the
generalized least squares (GLS) estimator, particularly under
compound symmetry.

## Design criteria and cost considerations

The formal framework for these results relies on classical
optimal design theory, as developed by @atkinson1992optimum.
@ouwens2002maximin proposed maximin D-optimal designs for
longitudinal mixed effects models, computing optimal allocations
of time-points for polynomial trend models with random
intercepts, random slopes, and AR(1) serial correlations. Because
D-optimal designs depend on unknown parameter values, the
maximin criterion provides robustness by maximizing the worst-case
relative efficiency across the parameter space.

@tekle2008doptimal extended this work to cohort designs, showing
that a purely longitudinal design with a single cohort measured
at optimal time-points is generally more efficient than mixed
longitudinal-cross-sectional designs. Their numerical results
demonstrated that the most efficient number of repeated
measurements equals the sum of the number of cohorts and the
degree of the polynomial trend. Subsequently,
@tekle2011cohorts showed that adding cohorts or repeated
measurements beyond this optimum wastes resources when the study
budget is fixed.

## Practical guidance on the number of measures

From a more pragmatic perspective, @vickers2003howmany examined
the marginal benefit of adding repeated measures in terms of
statistical power. He showed that while repeating assessments
can have dramatic effects on power, the marginal benefit of each
additional measure decreases rapidly. There is little value in
exceeding four post-treatment assessments (or seven total
including baseline), except when correlations between measures
are low, as in episodic conditions.

## Extensions and related work

Several extensions of the Winkens framework have appeared.
@winkens2007polynomial generalized the optimal design results to
second-order polynomial treatment effects, finding that equidistant
time-points are either optimal or highly efficient when only three
measures are taken. @candel2008empirical investigated optimal
designs when the goal is to estimate individual growth curves
(random effects) rather than fixed effects, showing that the
optimal number of time-points and subjects differs from the
fixed-effects case. @abebe2014bayesian developed Bayesian
D-optimal designs for dichotomous repeated measures with
autocorrelation, formally accounting for parameter uncertainty
through prior distributions.

For the analysis of unequally spaced longitudinal data,
@jones1991unequally provided the statistical machinery for
fitting mixed models with continuous-time AR(1) structures and
arbitrary observation schedules. Their approach, based on the
Kalman filter, remains foundational for analyzing data from
designs with non-equidistant time-points.

## Power and sample size tools

Modern software for power and sample size calculations in
longitudinal trials has made these design considerations more
accessible. @iddi2022longpower describe the `longpower` R
package, which implements closed-form power calculations for
linear mixed models. @donohue2010power applied these methods
to Alzheimer's disease Phase II trial design, demonstrating
the practical impact of design choices on the ability to detect
disease-modifying treatment effects.

## Present study

Despite the theoretical optimality results, there remains
limited simulation-based evidence illustrating how different
visit schedules perform in finite samples under realistic data
generating mechanisms. In this report, we conduct a simple
Monte Carlo study to compare two five-visit designs for a
100-patient, two-arm longitudinal trial with linearly divergent
treatment effects:

- **Equal spacing:** observations at months 0, 6, 12, 18, and 24.
- **Clustered spacing:** observations clustered near the
  beginning and end of the study period (months 0, 2, 4, 20, 24),
  motivated by the theoretical result that under compound
  symmetry, efficiency is gained by concentrating measures where
  the treatment effect is largest.

We evaluate the designs in terms of bias, standard error, power,
and coverage of 95% confidence intervals for the treatment-by-time
interaction.


# Methods

## Data generating model

We simulate data from a linear mixed effects model with random
intercepts and random slopes. Let $Y_{ij}$ denote the outcome for
subject $i$ at time $t_j$. The generating model is:

$$
Y_{ij} = (\beta_0 + b_{0i}) +
         (\beta_1 + b_{1i}) \, t_j +
         \beta_2 \, G_i +
         \beta_3 \, G_i \, t_j +
         \varepsilon_{ij}
$$

where $G_i \in \{0, 1\}$ is the treatment group indicator,
$b_{0i}$ and $b_{1i}$ are subject-specific random effects
with $\text{Var}(b_0) = \tau_0^2$, $\text{Var}(b_1) = \tau_1^2$,
$\text{Cov}(b_0, b_1) = \tau_{01}$, and
$\varepsilon_{ij} \sim N(0, \sigma^2)$.

The parameter of interest is $\beta_3$, the treatment-by-time
interaction, which represents the rate at which the treatment
effect diverges from the control effect per unit time.

## Parameter values

We use the following parameter values, broadly motivated by
cognitive decline trials:

```{r params}
n_subjects <- 100
n_per_group <- n_subjects / 2
beta_0 <- 50
beta_1 <- -1.0
beta_2 <- 0
beta_3 <- 0.5
tau_0 <- 5.0
tau_1 <- 0.3
tau_01 <- 0.0
sigma <- 3.0
n_sims <- 2000
study_end <- 24

times_equal <- c(0, 6, 12, 18, 24)
times_clustered <- c(0, 2, 4, 20, 24)

params <- data.frame(
  Parameter = c(
    "Intercept", "Time slope",
    "Group effect at baseline",
    "Group x time",
    "SD random intercept",
    "SD random slope",
    "Residual SD",
    "Subjects per group", "Study duration (months)",
    "Number of simulations"
  ),
  Symbol = c(
    "$\\beta_0$", "$\\beta_1$",
    "$\\beta_2$",
    "$\\beta_3$",
    "$\\tau_0$",
    "$\\tau_1$",
    "$\\sigma$",
    "$n$", "--",
    "--"
  ),
  Value = c(
    beta_0, beta_1, beta_2, beta_3,
    tau_0, tau_1, sigma,
    n_per_group, study_end, n_sims
  ),
  stringsAsFactors = FALSE
)

kable(
  params,
  format = "latex",
  escape = FALSE,
  booktabs = TRUE,
  caption = "Simulation parameters.",
  col.names = c("Parameter", "Symbol", "Value")
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

The treatment-by-time interaction $\beta_3 = 0.5$ implies that
the treatment group improves (or declines less) by 0.5 units per
month relative to control, yielding a cumulative treatment
advantage of `r beta_3 * study_end` units over `r study_end`
months.

## Design specifications

```{r design-table}
design_df <- tibble(
  Design = c("Equal", "Clustered"),
  `Visit 1` = c(times_equal[1], times_clustered[1]),
  `Visit 2` = c(times_equal[2], times_clustered[2]),
  `Visit 3` = c(times_equal[3], times_clustered[3]),
  `Visit 4` = c(times_equal[4], times_clustered[4]),
  `Visit 5` = c(times_equal[5], times_clustered[5])
)

kable(
  design_df,
  format = "latex",
  booktabs = TRUE,
  caption = "Observation times (months) for each design."
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

## Simulation procedure

For each of `r n_sims` Monte Carlo replications, we:

1. Generate `r n_subjects` subjects, randomized 1:1 to treatment
   and control.
2. Simulate longitudinal outcomes at the five visit times for each
   design using the data generating model above.
3. Fit a linear mixed model (random intercepts and slopes) using
   restricted maximum likelihood (REML) via `nlme::lme()`.
4. Extract the point estimate, standard error, and $p$-value for
   $\hat{\beta}_3$.

We then compute, for each design:

- Mean estimate of $\hat{\beta}_3$ (to assess bias).
- Empirical standard deviation of $\hat{\beta}_3$ (true SE).
- Mean model-based standard error.
- Empirical power (proportion of simulations with $p < 0.05$).
- Coverage of 95% confidence intervals.

```{r simulation, cache=TRUE}
simulate_trial <- function(times, n_per_group, beta_0,
                           beta_1, beta_2, beta_3,
                           tau_0, tau_1, tau_01, sigma) {
  n <- 2 * n_per_group
  k <- length(times)

  group <- rep(c(0, 1), each = n_per_group)
  id <- seq_len(n)

  b0 <- rnorm(n, 0, tau_0)
  b1 <- rnorm(n, 0, tau_1)

  dat <- expand.grid(visit = seq_len(k), id = id)
  dat$time <- times[dat$visit]
  dat$group <- group[dat$id]
  dat$b0 <- b0[dat$id]
  dat$b1 <- b1[dat$id]
  dat$y <- (beta_0 + dat$b0) +
    (beta_1 + dat$b1) * dat$time +
    beta_2 * dat$group +
    beta_3 * dat$group * dat$time +
    rnorm(nrow(dat), 0, sigma)

  dat$id <- factor(dat$id)
  dat
}

run_one_sim <- function(times, ...) {
  dat <- simulate_trial(times, ...)

  fit <- tryCatch(
    lme(
      y ~ time * group,
      random = ~ time | id,
      data = dat,
      method = "REML",
      control = lmeControl(
        opt = "optim",
        maxIter = 200,
        msMaxIter = 200
      )
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(
      data.frame(
        est = NA, se = NA, pval = NA,
        ci_lo = NA, ci_hi = NA
      )
    )
  }

  s <- summary(fit)
  coefs <- s$tTable
  row <- which(rownames(coefs) == "time:group")

  est <- coefs[row, "Value"]
  se <- coefs[row, "Std.Error"]
  pval <- coefs[row, "p-value"]
  df_res <- coefs[row, "DF"]
  ci_lo <- est - qt(0.975, df_res) * se
  ci_hi <- est + qt(0.975, df_res) * se

  data.frame(
    est = est, se = se, pval = pval,
    ci_lo = ci_lo, ci_hi = ci_hi
  )
}

sim_args <- list(
  n_per_group = n_per_group,
  beta_0 = beta_0, beta_1 = beta_1,
  beta_2 = beta_2, beta_3 = beta_3,
  tau_0 = tau_0, tau_1 = tau_1,
  tau_01 = tau_01, sigma = sigma
)

results_equal <- do.call(
  rbind,
  lapply(seq_len(n_sims), function(i) {
    do.call(run_one_sim, c(list(times = times_equal), sim_args))
  })
)
results_equal$design <- "Equal"

results_clustered <- do.call(
  rbind,
  lapply(seq_len(n_sims), function(i) {
    do.call(
      run_one_sim,
      c(list(times = times_clustered), sim_args)
    )
  })
)
results_clustered$design <- "Clustered"

results <- rbind(results_equal, results_clustered)
results <- results[!is.na(results$est), ]
```


# Results

## Summary statistics

```{r summary-table}
summary_df <- results |>
  group_by(design) |>
  summarise(
    n_converged = n(),
    mean_est = mean(est),
    bias = mean(est) - beta_3,
    empirical_se = sd(est),
    mean_model_se = mean(se),
    power = mean(pval < 0.05),
    coverage = mean(ci_lo <= beta_3 & beta_3 <= ci_hi),
    .groups = "drop"
  )

summary_out <- summary_df |>
  select(
    design, n_converged, mean_est, bias,
    empirical_se, mean_model_se, power, coverage
  )

kable(
  summary_out,
  format = "latex",
  booktabs = TRUE,
  digits = 4,
  col.names = c(
    "Design", "Converged", "Mean Est.",
    "Bias", "Emp. SE", "Model SE",
    "Power", "Coverage"
  ),
  caption = "Simulation results for the treatment-by-time
    interaction."
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 10
  )
```

```{r power-diff, results='asis'}
power_equal <- summary_df$power[summary_df$design == "Equal"]
power_clust <- summary_df$power[summary_df$design == "Clustered"]
se_equal <- summary_df$empirical_se[
  summary_df$design == "Equal"
]
se_clust <- summary_df$empirical_se[
  summary_df$design == "Clustered"
]
```

The equal-spacing design achieves an empirical power of
`r sprintf("%.1f%%", 100 * power_equal)`, compared with
`r sprintf("%.1f%%", 100 * power_clust)` for the clustered design.
The empirical standard error of $\hat{\beta}_3$ is
`r sprintf("%.4f", se_equal)` under equal spacing and
`r sprintf("%.4f", se_clust)` under the clustered design.

## Distribution of estimates

```{r density-plot, fig.cap="Sampling distribution of the treatment-by-time interaction estimate under each design.", fig.height=3.2}
ggplot(results, aes(x = est, linetype = design)) +
  geom_density(fill = NA, linewidth = 0.5) +
  geom_vline(
    xintercept = beta_3,
    linetype = "dotted",
    linewidth = 0.4
  ) +
  scale_linetype_manual(values = c(
    "Equal" = "solid",
    "Clustered" = "dashed"
  )) +
  labs(
    x = expression(hat(beta)[3]),
    y = "Density",
    linetype = "Design"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

## Standard errors across simulations

```{r se-boxplot, fig.cap="Distribution of model-based standard errors across simulations.", fig.height=3.0}
ggplot(results, aes(x = design, y = se)) +
  geom_boxplot(
    width = 0.5,
    outlier.size = 0.5,
    fill = "grey85"
  ) +
  labs(
    x = "Design",
    y = expression("SE of " ~ hat(beta)[3])
  ) +
  theme_minimal(base_size = 10)
```

## Design comparison

```{r design-visual, fig.cap="Schematic of the two visit schedules over the 24-month study period. Lines show the expected group trajectories.", fig.height=3.0}
time_grid <- seq(0, 24, length.out = 100)
traj_ctrl <- beta_0 + beta_1 * time_grid
traj_trt <- beta_0 + (beta_1 + beta_3) * time_grid

traj_df <- tibble(
  time = rep(time_grid, 2),
  y = c(traj_ctrl, traj_trt),
  Group = rep(c("Control", "Treatment"), each = 100)
)

visit_df <- tibble(
  time = c(times_equal, times_clustered),
  design = rep(
    c("Equal", "Clustered"), each = 5
  ),
  y_pos = c(
    rep(beta_0 + beta_1 * 12 + 8, 5),
    rep(beta_0 + beta_1 * 12 + 5, 5)
  )
)

ggplot() +
  geom_line(
    data = traj_df,
    aes(x = time, y = y, linetype = Group),
    linewidth = 0.5
  ) +
  geom_point(
    data = visit_df,
    aes(x = time, y = y_pos, shape = design),
    size = 2.5
  ) +
  scale_linetype_manual(values = c(
    "Control" = "solid",
    "Treatment" = "dashed"
  )) +
  scale_shape_manual(values = c(
    "Equal" = 16,
    "Clustered" = 17
  )) +
  labs(
    x = "Month",
    y = "Expected outcome",
    shape = "Design",
    linetype = "Group"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```


# Discussion

This simulation study illustrates the practical consequences of
visit placement in longitudinal clinical trials. Both designs use
the same number of subjects and the same number of visits; they
differ only in the timing of observations.

The results are broadly consistent with the theoretical findings
of @winkensOptimalTimepointsClinical2005. When the data generating
mechanism includes both random slopes and independent residual
error -- a structure that combines features of compound symmetry
and AR(1) with measurement error -- the clustered design, which
concentrates observations near the boundaries of the study period,
can yield different precision for the treatment-by-time interaction
compared with equal spacing.

Several caveats apply. First, we assumed complete data with no
dropout. In practice, attrition is common in longitudinal trials
and may differentially affect designs that place observations at
extreme time-points. Second, the data generating model used
independent random effects and homoscedastic residuals; more
complex covariance structures (e.g., heterogeneous variances,
serial correlation in residuals) could alter the relative
performance of the two designs. Third, we considered only a
linear treatment effect; @winkens2007polynomial showed that
optimal designs differ for polynomial effects of higher order.

Future work should investigate the robustness of these findings
to dropout, misspecification of the covariance structure, and
nonlinear treatment trajectories. The framework of
@ouwens2002maximin for maximin D-optimal designs provides one
principled approach to robust design when the covariance structure
is uncertain. Bayesian approaches, as in @abebe2014bayesian,
offer another avenue for incorporating prior uncertainty about
model parameters into the design optimization.


# References


::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
