---
title: "Efficiency Consequences of Visit Schedule Choice in
  Longitudinal Trials: An Analytic and Simulation-Based Comparison"
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
abstract: |
  Observation times in longitudinal clinical trials are usually
  spaced equally for administrative convenience, but optimal design
  theory shows that placement affects the precision of estimated
  treatment effects. We compare an equally spaced five-visit
  schedule with a boundary-clustered schedule for a two-arm,
  24-month trial with a linearly divergent treatment effect,
  combining an exact generalized least squares (GLS) efficiency
  calculation with a common-random-number Monte Carlo study (2000
  paired replications) under a random-intercept, random-slope data
  generating model. The analytic relative efficiency of the
  clustered design is 1.06, and the paired simulation, after
  recalibrating the effect size so power is informative rather than
  at ceiling, reproduces this result (empirical relative efficiency
  1.07, 95% bootstrap interval 1.06-1.09), confirming that the
  precision advantage is real and not simulation noise. A subsequent
  Monte Carlo grid crossing five covariance structures with four
  visit schedules corroborates the direction of this finding in 14
  of 15 non-baseline cells, with the advantage shrinking as the
  random-slope variance grows large relative to the residual
  variance. We do not claim optimality: the theoretically optimal
  design for this covariance structure is not computed here, and the
  primary paired comparison uses only two designs. The contribution
  is a worked, fully reproducible demonstration of how much
  efficiency is at stake for one realistic configuration, and a
  template (`R/simulation.R`) for extending the comparison to
  dropout mechanisms and further candidate schedules.
keywords: "longitudinal data; optimal design; visit schedule;
  linear mixed model; Monte Carlo simulation; relative efficiency"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
library(nlme)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

# Load the visitplacement package functions (simulate_trial(),
# run_one_sim(), analytic_gls_se(), analytic_gls_se_general(),
# cov_*()) from R/ rather than defining them inline, per the
# 2026-08-16 review (issue 2.8). pkgload::load_all() is used
# instead of library(visitplacement) because the package is not
# installed; it works whether the report is knit from within the
# package directory or from analysis/report/ via the here-relative
# project root.
pkg_root <- rprojroot::find_root(
  rprojroot::has_file("DESCRIPTION"),
  path = here::here()
)
pkgload::load_all(pkg_root, quiet = TRUE)

set.seed(20260305)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5.5,
  fig.height = 3.5,
  fig.align = "center",
  dev = "pdf"
)

options(scipen = 1, digits = 3)
```

# Introduction

In longitudinal clinical trials, subjects are measured
repeatedly over a fixed study period. The conventional approach
assigns observation times at equally spaced intervals (monthly
or quarterly visits, for example) largely for administrative
convenience rather than statistical efficiency. A growing body
of work in optimal experimental design suggests that the
placement of these observation times can substantially affect
the precision with which treatment effects are estimated. We
begin in Section \@ref(optimal-allocation-of-observation-times)
with a review of the relevant literature and then describe the
analytic efficiency calculation and simulation study, and their
results, in Sections \@ref(methods) and \@ref(results).

## Optimal allocation of observation times

@winkensOptimalTimepointsClinical2005 provided the foundational
analysis of this problem for two-group trials with linearly
divergent treatment effects. They showed that the optimal
allocation of time-points depends critically on the assumed
covariance structure. Under compound symmetry (CS), efficiency
improves by clustering repeated measures near the end of the
study period, where the treatment effect is largest. Under a
first-order autoregressive (AR(1)) structure, only two
measurements -- at baseline and at study end -- are needed for
highly efficient estimation. Under a more realistic AR(1) plus
measurement error structure, the optimal spacing depends on the
specific covariance parameter values, but the equally spaced
design is generally suboptimal.

In a companion paper, @winkensOptimalNumberRepeated2006 extended
these results to address the joint optimization of the number
of repeated measures and group sizes under budget constraints.
They found that when budget is fixed and the covariance structure
includes an autoregressive component, two repeated measures per
subject often yield highly efficient treatment effect estimators.
Otherwise, intermediate measures can reduce the variance of the
generalized least squares (GLS) estimator, particularly under
compound symmetry.

## Design criteria and cost considerations

The formal framework for these results relies on classical
optimal design theory, as developed by @atkinson1992optimum.
@ouwens2002maximin proposed maximin D-optimal designs for
longitudinal mixed effects models, computing optimal allocations
of time-points for polynomial trend models with random
intercepts, random slopes, and AR(1) serial correlations. Because
D-optimal designs depend on unknown parameter values, the
maximin criterion provides robustness by maximizing the worst-case
relative efficiency across the parameter space.

@tekle2008doptimal extended this work to cohort designs, showing
that a purely longitudinal design with a single cohort measured
at optimal time-points is generally more efficient than mixed
longitudinal-cross-sectional designs. Their numerical results
demonstrated that the most efficient number of repeated
measurements equals the sum of the number of cohorts and the
degree of the polynomial trend. Subsequently,
@tekle2011cohorts showed that adding cohorts or repeated
measurements beyond this optimum wastes resources when the study
budget is fixed.

## Practical guidance on the number of measures

From a more pragmatic perspective, @vickers2003howmany examined
the marginal benefit of adding repeated measures in terms of
statistical power. He showed that while repeating assessments
can have dramatic effects on power, the marginal benefit of each
additional measure decreases rapidly. In general, there is little value in exceeding four
post-treatment assessments (or seven total including baseline),
except when correlations between measures are low, as in
episodic conditions.

## Extensions and related work

Several extensions of the Winkens framework have appeared.
@winkens2007polynomial generalized the optimal design results to
second-order polynomial treatment effects, finding that equidistant
time-points are either optimal or highly efficient when only three
measures are taken. @candel2008empirical investigated optimal
designs when the goal is to estimate individual growth curves
(random effects) rather than fixed effects, showing that the
optimal number of time-points and subjects differs from the
fixed-effects case. @abebe2014bayesian developed Bayesian
D-optimal designs for dichotomous repeated measures with
autocorrelation, formally accounting for parameter uncertainty
through prior distributions.

For the analysis of unequally spaced longitudinal data,
@jones1991unequally provided the statistical machinery for
fitting mixed models with continuous-time AR(1) structures and
arbitrary observation schedules. Their approach, based on the
Kalman filter, remains foundational for analyzing data from
designs with non-equidistant time-points.

## Power and sample size tools

Modern software for power and sample size calculations in
longitudinal trials has made these design considerations more
accessible. @iddi2022longpower describe the `longpower` R
package, which implements closed-form power calculations for
linear mixed models. @donohue2010power applied these methods
to Alzheimer's disease Phase II trial design, demonstrating
the practical impact of design choices on the ability to detect
disease-modifying treatment effects.

## Present study

The theoretical optimality results reviewed above establish that
the best design depends on the assumed covariance structure, but
they do not by themselves quantify how much efficiency is at
stake for a specific, realistic trial configuration, nor do they
address the finite-sample behavior of the REML-based inference
that practitioners actually use. This report addresses that
translational gap for one configuration: a 100-patient, two-arm,
24-month trial with a linearly divergent treatment effect,
analyzed under a random-intercept, random-slope model. We compare
two five-visit designs:

- **Equal spacing:** observations at months 0, 6, 12, 18, and 24.
- **Clustered spacing:** observations clustered near the
  beginning and end of the study period (months 0, 2, 4, 20, 24).
  Under a random-slope covariance structure (Section
  \@ref(data-generating-model)), boundary-loaded time-points are
  efficient for estimating the slope contrast for a classical
  leverage reason -- observations far from the mean time carry more
  information about a slope -- which is distinct from the
  compound-symmetry mechanism that motivates clustering in
  @winkensOptimalTimepointsClinical2005; see Section
  \@ref(covariance-structure-and-theoretical-prediction) for the
  derivation.

We first derive the exact asymptotic (GLS) relative efficiency of
the two designs in closed form, then use a common-random-number
Monte Carlo study to assess whether finite-sample REML estimation
and inference (bias, empirical versus model-based standard error,
power, and 95% confidence interval coverage) reproduce the
analytic prediction. We do not claim that either design is
optimal: the theoretically optimal design implied by the cited
literature is not computed here, and the main paired comparison
considers only two candidate schedules, although a Monte Carlo
sensitivity grid over four schedules and five covariance structures
(Section \@ref(scenario-extension-monte-carlo)) corroborates the
direction of the finding more broadly. `R/simulation.R` in the
accompanying compendium implements the data-generating model, the
analytic efficiency calculation, and the paired simulation as
reusable functions so the comparison can be extended to other
covariance structures, sample sizes, dropout mechanisms, and
candidate schedules (Sections
\@ref(scenario-extension-analytic-only) and
\@ref(scenario-extension-monte-carlo)).


# Methods

## Data generating model

We simulate data from a linear mixed effects model with random
intercepts and random slopes. Let $Y_{ij}$ denote the outcome for
subject $i$ at time $t_j$. The generating model is:

$$
Y_{ij} = (\beta_0 + b_{0i}) +
         (\beta_1 + b_{1i}) \, t_j +
         \beta_2 \, G_i +
         \beta_3 \, G_i \, t_j +
         \varepsilon_{ij}
$$

where $G_i \in \{0, 1\}$ is the treatment group indicator,
$b_{0i}$ and $b_{1i}$ are subject-specific random effects
with $\text{Var}(b_0) = \tau_0^2$, $\text{Var}(b_1) = \tau_1^2$,
$\text{Cov}(b_0, b_1) = \tau_{01}$, and
$\varepsilon_{ij} \sim N(0, \sigma^2)$.

The parameter of interest is $\beta_3$, the treatment-by-time
interaction, which represents the rate at which the treatment
effect diverges from the control effect per unit time.

## Covariance structure and theoretical prediction

Marginalizing over the random effects, this model induces a
within-subject covariance
$\text{Cov}(Y_{ij}, Y_{ij'}) = \tau_0^2 + \tau_{01}(t_j + t_{j'}) +
\tau_1^2 t_j t_{j'} + \sigma^2 \mathbb{1}(j = j')$, in which the
marginal variance increases with $t_j^2$ and the correlation
between visits depends on their product $t_j t_{j'}$, not on their
lag $|t_j - t_{j'}|$. This is the standard random-coefficients
(random-intercept, random-slope) covariance structure. It is
distinct from both compound symmetry (CS), which is the
random-intercept-only special case ($\tau_1 = 0$) with constant
correlation regardless of visit timing, and from a first-order
autoregressive (AR(1)) structure, whose correlation depends only on
the lag between visits and does not grow with $\tau_1$ boundary
leverage. An earlier draft of this report described the simulated
structure as combining features of CS and AR(1); that
characterization is incorrect and has been removed (see the
Discussion for the corrected interpretation).

The efficiency prediction relevant to this structure differs from
the CS-based argument in @winkensOptimalTimepointsClinical2005.
Under CS, clustering observations near the end of the study helps
because the treatment effect (and hence the signal) is largest
there while every pair of visits is equally correlated, so nothing
is lost by leaving a gap. Under the random-slope structure
simulated here, the reason boundary-clustered designs are more
efficient for the *slope* contrast $\beta_3$ is instead the
classical leverage argument for regression slopes: for a simple
linear regression of $Y$ on $t$, $\text{Var}(\hat{\text{slope}})
\propto 1 / \sum_i (t_i - \bar{t})^2$, so observations placed far
from the mean time carry disproportionately more information about
the slope. Section \@ref(analytic-relative-efficiency) below
verifies this prediction is directionally correct but shows the
efficiency gain shrinks as the random-slope variance $\tau_1^2$
grows relative to the residual variance, because a larger
$\tau_1^2$ makes the slope itself (not just its estimation error)
more variable across subjects, which increases the total variance
contributed by every visit and dilutes the boundary-leverage
advantage.

## Parameter values

We use the following parameter values, broadly motivated by
cognitive decline trials:

```{r params}
n_subjects <- 100
n_per_group <- n_subjects / 2
beta_0 <- 50
beta_1 <- -1.0
beta_2 <- 0
beta_3 <- 0.19
tau_0 <- 5.0
tau_1 <- 0.3
tau_01 <- 0.0
sigma <- 3.0
n_sims <- 2000
study_end <- 24

times_equal <- c(0, 6, 12, 18, 24)
times_clustered <- c(0, 2, 4, 20, 24)

params <- data.frame(
  Parameter = c(
    "Intercept", "Time slope",
    "Group effect at baseline",
    "Group x time",
    "SD random intercept",
    "SD random slope",
    "Residual SD",
    "Subjects per group", "Study duration (months)",
    "Number of simulations"
  ),
  Symbol = c(
    "$\\beta_0$", "$\\beta_1$",
    "$\\beta_2$",
    "$\\beta_3$",
    "$\\tau_0$",
    "$\\tau_1$",
    "$\\sigma$",
    "$n$", "--",
    "--"
  ),
  Value = c(
    beta_0, beta_1, beta_2, beta_3,
    tau_0, tau_1, sigma,
    n_per_group, study_end, n_sims
  ),
  stringsAsFactors = FALSE
)

kable(
  params,
  format = "latex",
  escape = FALSE,
  booktabs = TRUE,
  caption = "Simulation parameters.",
  col.names = c("Parameter", "Symbol", "Value")
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

The treatment-by-time interaction $\beta_3 = 0.19$ implies that
the treatment group improves (or declines less) by 0.19 units per
month relative to control, yielding a cumulative treatment
advantage of `r beta_3 * study_end` units over `r study_end`
months.

### Recalibration of the treatment effect

An earlier draft used $\beta_3 = 0.5$, for which the analytic power
of both designs is 1 to machine precision (Section
\@ref(analytic-relative-efficiency)): every replication rejects the
null, so power cannot discriminate between designs and the metric
is uninformative. We recalibrated $\beta_3$ so that the
equal-spacing design has approximately 80% analytic power at the
two-sided $\alpha = 0.05$ level, which requires
$\beta_3 / \text{SE}_{\text{equal}} \approx 2.80$ (the sum of the
$z$-quantiles for 80% power and a two-sided 5% test). With
$\text{SE}_{\text{equal}} = 0.0678$ (Section
\@ref(analytic-relative-efficiency)), this gives $\beta_3 \approx
0.19$, for which the analytic power is 80.0% (equal spacing) and
82.4% (clustered spacing) -- close to, and not at, the ceiling, so
that power remains informative while both designs are still
adequately powered for a realistic trial. All other parameters are
unchanged from the original design.

## Design specifications

```{r design-table}
design_df <- tibble(
  Design = c("Equal", "Clustered"),
  `Visit 1` = c(times_equal[1], times_clustered[1]),
  `Visit 2` = c(times_equal[2], times_clustered[2]),
  `Visit 3` = c(times_equal[3], times_clustered[3]),
  `Visit 4` = c(times_equal[4], times_clustered[4]),
  `Visit 5` = c(times_equal[5], times_clustered[5])
)

kable(
  design_df,
  format = "latex",
  booktabs = TRUE,
  caption = "Observation times (months) for each design."
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

```{r design-visual, fig.cap="Schematic of the two visit schedules over the 24-month study period. Lines show the expected group trajectories; markers show each design's visit times, vertically offset (Equal above, Clustered below) purely to avoid overplotting at shared visit times (0 and 24 months) and carry no other meaning.", fig.height=3.0}
time_grid <- seq(0, 24, length.out = 100)
traj_ctrl <- beta_0 + beta_1 * time_grid
traj_trt <- beta_0 + (beta_1 + beta_3) * time_grid

traj_df <- tibble(
  time = rep(time_grid, 2),
  y = c(traj_ctrl, traj_trt),
  Group = rep(c("Control", "Treatment"), each = 100)
)

visit_df <- tibble(
  time = c(times_equal, times_clustered),
  design = rep(
    c("Equal", "Clustered"), each = 5
  ),
  y_pos = c(
    rep(beta_0 + beta_1 * 12 + 3, 5),
    rep(beta_0 + beta_1 * 12 + 1, 5)
  )
)

ggplot() +
  geom_line(
    data = traj_df,
    aes(x = time, y = y, linetype = Group),
    linewidth = 0.5
  ) +
  geom_point(
    data = visit_df,
    aes(x = time, y = y_pos, shape = design),
    size = 2.5
  ) +
  scale_linetype_manual(values = c(
    "Control" = "solid",
    "Treatment" = "dashed"
  )) +
  scale_shape_manual(values = c(
    "Equal" = 16,
    "Clustered" = 17
  )) +
  labs(
    x = "Month",
    y = "Expected outcome",
    shape = "Design",
    linetype = "Group"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

## Analytic relative efficiency

Under the data generating model of Section
\@ref(data-generating-model), the treatment-by-time interaction
$\hat{\beta}_3$ has an exact large-sample GLS variance: with
$\mathbf{Z}_i = [\mathbf{1}, \mathbf{t}]$ the $k \times 2$ per-subject
design matrix for the random effects, $\mathbf{D}$ the random-effects
covariance matrix, and
$\mathbf{V} = \mathbf{Z}_i \mathbf{D} \mathbf{Z}_i' + \sigma^2
\mathbf{I}_k$, the information matrix
$\mathbf{M} = \sum_i \mathbf{X}_i' \mathbf{V}^{-1} \mathbf{X}_i$
(summed separately over control subjects, for whom the last two
columns of $\mathbf{X}_i$ are zero, and treatment subjects) gives
$\text{Var}(\hat{\beta}_3) = [\mathbf{M}^{-1}]_{44}$. This ten-line
linear algebra calculation, implemented as `analytic_gls_se()` in
`R/simulation.R`, requires no simulation:

```{r analytic-efficiency}
analytic_se_equal <- analytic_gls_se(
  times_equal, n_per_group, tau_0, tau_1, tau_01, sigma
)
analytic_se_clustered <- analytic_gls_se(
  times_clustered, n_per_group, tau_0, tau_1, tau_01, sigma
)
analytic_re <- (analytic_se_equal / analytic_se_clustered)^2

analytic_tbl <- tibble(
  Design = c("Equal", "Clustered"),
  `Analytic SE` = c(analytic_se_equal, analytic_se_clustered)
)

kable(
  analytic_tbl,
  format = "latex",
  booktabs = TRUE,
  digits = 4,
  caption = paste(
    "Exact GLS standard errors of the treatment-by-time",
    "interaction under the random-slope data-generating model."
  )
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 10
  )
```

The exact GLS standard errors are
`r sprintf("%.4f", analytic_se_equal)` (equal spacing) and
`r sprintf("%.4f", analytic_se_clustered)` (clustered spacing),
giving a relative efficiency (ratio of variances) of
`r sprintf("%.3f", analytic_re)` in favor of the clustered design.
This analytic value is the primary quantity of interest for the
design comparison; the simulation in Section
\@ref(simulation-procedure) exists to answer a different question --
whether finite-sample REML estimation and Wald-type inference
(bias, coverage, and empirical versus model-based SE) reproduce
this analytic prediction and behave as expected at $n = 100$ -- not
to re-derive a number already available in closed form.

## Scenario extension (analytic only)

The single scenario simulated in this report fixes one covariance
structure, one effect size, and one sample size (2026-08-16
review, issue 2.4). Because the design comparison's quantity of
interest, the GLS relative efficiency, has a closed form (Section
\@ref(analytic-relative-efficiency)), `analytic_gls_se_general()`
in `R/covariance.R` generalizes the calculation to an arbitrary
within-subject covariance matrix, so the sensitivity of the
efficiency contrast to the covariance structure can be assessed
without additional Monte Carlo simulation. Table
\@ref(tab:scenario-grid) reports the exact relative efficiency of
the clustered versus equal-spacing design across the random-slope
SD $\tau_1$ used in the main simulation, and for two additional
structures: compound symmetry, the mechanism that motivates
clustering in @winkensOptimalTimepointsClinical2005, and AR(1)
plus measurement error.

```{r scenario-grid}
scenario_grid <- readRDS(
  here::here(
    "analysis", "data", "derived_data", "scenario_grid.rds"
  )
)

kable(
  scenario_grid,
  format = "latex",
  booktabs = TRUE,
  digits = 4,
  col.names = c(
    "Structure", "Parameter", "SE (Equal)",
    "SE (Clustered)", "Relative efficiency"
  ),
  caption = paste(
    "Analytic (closed-form, no simulation) relative efficiency of",
    "the clustered versus equal-spacing design across covariance",
    "structures and random-slope variance."
  )
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 9
  )
```

The clustered design is more efficient across the full
random-slope grid, but the advantage shrinks as $\tau_1$ grows
relative to the residual SD $\sigma$, consistent with the
leverage-dilution argument of Section
\@ref(covariance-structure-and-theoretical-prediction). Under
compound symmetry and under AR(1) plus measurement error the
clustered design is also more efficient, although by a different
mechanism in each case. This extension is analytic only: it does
not assess whether finite-sample REML estimation and inference
reproduce these predictions, nor does it vary sample size or
dropout. Section \@ref(scenario-extension-monte-carlo) extends the
comparison to a genuine Monte Carlo grid over covariance structure
and visit schedule; dropout remains future work (see the
Discussion). `Rscript analysis/scripts/run_scenario_grid.R`
regenerates Table \@ref(tab:scenario-grid).

## Scenario extension (Monte Carlo)

```{r load-factorial-grid}
factorial_grid <- readRDS(
  here::here(
    "analysis", "data", "derived_data", "factorial_grid.rds"
  )
)
factorial_n_sims <- factorial_grid$params$n_sims
```

The preceding section is analytic only: it evaluates the
closed-form asymptotic GLS efficiency but does not simulate
finite-sample data or fit REML models away from the single
random-slope design actually used in Section
\@ref(simulation-procedure). This section extends the Monte Carlo
comparison itself to a factorial grid crossing covariance
structure (the random-slope SD $\tau_1$ grid, compound symmetry,
and AR(1) plus measurement error, matching Table
\@ref(tab:scenario-grid)) with four candidate five-visit schedules
spanning a range of boundary clustering: equal spacing (months 0,
6, 12, 18, 24, as in the main comparison), mild clustering (0, 4,
8, 20, 24), boundary clustering (0, 2, 4, 20, 24, the "Clustered"
design of the main comparison), and extreme clustering (0, 1, 2,
23, 24). For each of the resulting 20 cells, `run_one_sim()` (for
the random-slope structures) or `run_one_sim_cov()` (for compound
symmetry and AR(1) plus error, fitting `nlme::gls()` with a
matched correlation structure) is used to fit
`r factorial_n_sims` simulated replications under REML, exactly as
in the main comparison. This is a genuine Monte Carlo assessment,
not a re-derivation of the closed-form values, and it explicitly
excludes dropout: adding a dropout mechanism requires a design
decision (MCAR/MAR, rate, whether attrition is tied to visit
timing) left to the author (see the Discussion). Replication count
per cell (`r factorial_n_sims`) is reduced from the main
comparison's `r n_sims` to keep the 20-cell grid within a practical
single-core runtime; the completed grid took
`r sprintf("%.0f", factorial_grid$run_info$elapsed_seconds / 60)`
minutes. `Rscript analysis/scripts/run_full_factorial.R`
regenerates Table \@ref(tab:factorial-grid).

```{r factorial-grid}
factorial_tbl <- factorial_grid$grid[, c(
  "structure", "parameter", "schedule", "power", "empirical_se",
  "relative_efficiency_vs_equal"
)]

kable(
  factorial_tbl,
  format = "latex",
  booktabs = TRUE,
  digits = c(0, 0, 0, 3, 4, 3),
  col.names = c(
    "Structure", "Parameter", "Schedule", "Power", "Emp. SE",
    "Rel. eff. vs. Equal"
  ),
  caption = paste(
    "Monte Carlo relative efficiency of each visit schedule versus",
    "equal spacing, by covariance structure, at",
    factorial_n_sims, "replications per cell."
  )
) |>
  kable_styling(
    latex_options = "hold_position",
    font_size = 8
  )
```

```{r factorial-highlights}
fg <- factorial_grid$grid
max_re <- max(fg$relative_efficiency_vs_equal, na.rm = TRUE)
se_boundary_tau50 <- fg$empirical_se[
  fg$structure == "Random slope" & fg$parameter == "tau_1 = 0.50" &
    fg$schedule == "Boundary-clustered"
]
se_equal_tau50 <- fg$empirical_se[
  fg$structure == "Random slope" & fg$parameter == "tau_1 = 0.50" &
    fg$schedule == "Equal"
]
re_boundary_tau50 <- fg$relative_efficiency_vs_equal[
  fg$structure == "Random slope" & fg$parameter == "tau_1 = 0.50" &
    fg$schedule == "Boundary-clustered"
]
re_ar1_extreme <- fg$relative_efficiency_vs_equal[
  fg$structure == "AR(1) + error" & fg$schedule == "Extreme-clustered"
]
re_ar1_boundary <- fg$relative_efficiency_vs_equal[
  fg$structure == "AR(1) + error" & fg$schedule == "Boundary-clustered"
]
```

Across 14 of the 15 non-baseline (structure, schedule) cells,
moving from equal to mild-, boundary-, or extreme-clustered spacing
increases empirical precision (relative efficiency vs. equal
spacing exceeds 1), corroborating -- with
simulated finite-sample REML/GLS fits rather than the closed-form
calculation alone -- the direction of the main comparison's finding
across every covariance structure examined. The clustering
advantage is largest under compound symmetry and the small-$\tau_1$
random-slope structure (relative efficiency up to
`r sprintf("%.2f", max_re)` for extreme clustering), consistent
with Table \@ref(tab:scenario-grid)'s analytic prediction that the
advantage is strongest when the random-slope variance is small
relative to the residual variance. At $\tau_1 = 0.50$, where that
variance ratio is largest, the pattern is no longer monotonic and
the boundary-clustered design's empirical SE
(`r sprintf("%.4f", se_boundary_tau50)`) is slightly *larger* than
equal spacing's (`r sprintf("%.4f", se_equal_tau50)`, relative
efficiency `r sprintf("%.2f", re_boundary_tau50)`); given each
cell's Monte Carlo standard error of the empirical SE (on the order
of 0.002 at this sample size, from the same formula as Section
\@ref(simulation-procedure)), this single reversal is consistent
with Monte Carlo noise around a true relative efficiency close to 1
rather than a genuine crossover, but it underscores that the
clustering advantage is not guaranteed to hold once the
random-slope variance is large relative to the residual variance --
precisely the boundary case the analytic grid already flagged. The
AR(1)-plus-error structure shows a smaller, non-monotonic advantage
(extreme clustering is less efficient than boundary clustering,
relative efficiency `r sprintf("%.2f", re_ar1_extreme)` vs.
`r sprintf("%.2f", re_ar1_boundary)`), plausibly because packing
visits very close together adds
little independent information once residuals are themselves
serially correlated at short lags. All 20 cells converged fully
(`r min(factorial_grid$grid$n_converged)`/`r factorial_n_sims`
replications at minimum); estimated bias is negligible in every
cell (magnitude below 0.004, smaller than its own MCSE in most
cells), so REML estimation of the treatment-by-time interaction
is not detectably biased by covariance misspecification or
boundary clustering at this sample size.

## Simulation procedure

```{r load-sim-results}
sim_results <- readRDS(
  here::here("analysis", "data", "derived_data", "sim_results.rds")
)
results <- rbind(
  transform(sim_results$raw$equal, design = "Equal"),
  transform(sim_results$raw$clustered, design = "Clustered")
)
results <- results[!is.na(results$est), ]
```

The simulation functions (`simulate_trial()`, `run_one_sim()`,
`run_paired_simulation()`, `analytic_gls_se()`) live in `R/` as
documented, unit-tested package functions (`inst/tinytest/`) rather
than inline in this document, so they can be reused for the
scenario extension and audited independently of the manuscript
(2026-08-16 review, issue 2.8). For each of `r n_sims` Monte Carlo
replications, `run_paired_simulation()`:

1. Generates `r n_subjects` subjects, randomized 1:1 to treatment
   and control, drawing one shared set of subject-level random
   effects $(b_{0i}, b_{1i})$ and one shared set of residual errors
   $\varepsilon_{ij}$ per subject-visit index $j = 1, \dots, 5$.
2. Builds the equal-spacing and clustered-spacing datasets from
   *those same draws*, so the $j$-th random effect and residual
   realization is reused across designs for the $j$-th visit rather
   than redrawn -- these are common random numbers (CRN), which
   pair the two designs within each replication and substantially
   reduce the Monte Carlo error of the *difference* between designs
   relative to two independent simulation arms.
3. Fits a linear mixed model (random intercepts and slopes) to each
   design's dataset using restricted maximum likelihood (REML) via
   `nlme::lme()`.
4. Extracts the point estimate, standard error, and $p$-value for
   $\hat{\beta}_3$ for each design.

For each design we compute the mean estimate (bias), the empirical
SD of $\hat{\beta}_3$, the mean model-based SE, empirical power
($p < 0.05$), and 95% CI coverage, each with a Monte Carlo standard
error (MCSE) computed following @morris2019simulation: MCSE of the
empirical SD is $\text{SE} / \sqrt{2(R-1)}$, MCSE of a proportion
(power, coverage) is $\sqrt{\hat p (1-\hat p) / R}$, and MCSE of the
bias is $\text{SE} / \sqrt{R}$, where $R$ is the number of converged
replications. We additionally report a paired bootstrap (resampling
replication indices, which respects the CRN pairing) 95% interval
for the relative efficiency (ratio of empirical variances).

The driver script `analysis/scripts/run_simulation.R` runs this
procedure and saves the raw per-replication estimates, the summary
statistics with MCSEs, the analytic SEs, and the bootstrap interval
to a single versioned artifact,
`analysis/data/derived_data/sim_results.rds`, which this document
reads directly rather than re-simulating on every render or relying
on the knitr cache to stay synchronized with the reported numbers
(2026-08-16 review, issue 2.8(e)). To regenerate:
`Rscript analysis/scripts/run_simulation.R 2000`. The results below
used `r n_sims` replications, `r sim_results$run_info$r_version`,
and `nlme` `r sim_results$run_info$nlme_version`, completing in
`r sprintf("%.0f", sim_results$run_info$elapsed_seconds)` seconds
(single core, no parallelization).


# Results

## Summary statistics

```{r summary-table}
summary_df <- sim_results$summary

summary_out <- summary_df |>
  select(
    design, n_converged, mean_est, bias, mcse_bias,
    empirical_se, mcse_empirical_se, mean_model_se,
    power, mcse_power, coverage, mcse_coverage
  )

kable(
  summary_out,
  format = "latex",
  booktabs = TRUE,
  digits = 4,
  col.names = c(
    "Design", "Converged", "Mean Est.", "Bias",
    "MCSE(Bias)", "Emp. SE", "MCSE(Emp. SE)", "Model SE",
    "Power", "MCSE(Power)", "Coverage", "MCSE(Coverage)"
  ),
  caption = paste(
    "Simulation results for the treatment-by-time interaction,",
    "with Monte Carlo standard errors (MCSE) of each metric",
    "computed following Morris, White, and Crowther (2019)."
  )
) |>
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  )
```

Replications for which `nlme::lme()` failed to converge (an
error or warning during fitting, per `run_one_sim()` in
`R/simulation.R`) are recorded as `NA` and excluded from the
summary statistics above rather than imputed or retried; the
`Converged` column reports how many of the `r n_sims` attempted
replications per design contributed to each summary statistic.

```{r power-diff, results='asis'}
power_equal <- summary_df$power[summary_df$design == "Equal"]
mcse_power_equal <- summary_df$mcse_power[
  summary_df$design == "Equal"
]
power_clust <- summary_df$power[summary_df$design == "Clustered"]
mcse_power_clust <- summary_df$mcse_power[
  summary_df$design == "Clustered"
]
se_equal <- summary_df$empirical_se[summary_df$design == "Equal"]
mcse_se_equal <- summary_df$mcse_empirical_se[
  summary_df$design == "Equal"
]
se_clust <- summary_df$empirical_se[
  summary_df$design == "Clustered"
]
mcse_se_clust <- summary_df$mcse_empirical_se[
  summary_df$design == "Clustered"
]
re_boot <- sim_results$relative_efficiency_boot
```

We observe that the equal-spacing design achieves an empirical
power of `r sprintf("%.1f%% (MCSE %.1f%%)", 100 * power_equal,
100 * mcse_power_equal)`, compared with `r sprintf("%.1f%% (MCSE
%.1f%%)", 100 * power_clust, 100 * mcse_power_clust)` for the
clustered design; both are informative rather than at ceiling
following the recalibration of $\beta_3$ (Section
\@ref(recalibration-of-the-treatment-effect)). The empirical
standard error of $\hat{\beta}_3$ is `r sprintf("%.4f (MCSE
%.5f)", se_equal, mcse_se_equal)` under equal spacing and
`r sprintf("%.4f (MCSE %.5f)", se_clust, mcse_se_clust)` under
the clustered design. The paired-bootstrap relative efficiency (ratio
of empirical variances, resampling replication indices so the
common-random-number pairing is respected) is `r sprintf("%.2f",
re_boot$point)`, 95% bootstrap interval `r sprintf("%.2f",
re_boot$ci[1])`-`r sprintf("%.2f", re_boot$ci[2])`, consistent
with the analytic relative efficiency of `r sprintf("%.3f",
analytic_re)` (Section \@ref(analytic-relative-efficiency)); the
interval excludes 1, so the precision advantage of the clustered
design is distinguishable from Monte Carlo noise.

## Distribution of estimates

```{r density-plot, fig.cap="Sampling distribution of the treatment-by-time interaction estimate under each design.", fig.height=3.2}
ggplot(results, aes(x = est, linetype = design)) +
  geom_density(fill = NA, linewidth = 0.5) +
  geom_vline(
    xintercept = beta_3,
    linetype = "dotted",
    linewidth = 0.4
  ) +
  scale_linetype_manual(values = c(
    "Equal" = "solid",
    "Clustered" = "dashed"
  )) +
  labs(
    x = expression(hat(beta)[3]),
    y = "Density",
    linetype = "Design"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

## Standard errors across simulations

```{r se-boxplot, fig.cap="Distribution of model-based standard errors across simulations.", fig.height=3.0}
ggplot(results, aes(x = design, y = se)) +
  geom_boxplot(
    width = 0.5,
    outlier.size = 0.5,
    fill = "grey85"
  ) +
  labs(
    x = "Design",
    y = expression("SE of " ~ hat(beta)[3])
  ) +
  theme_minimal(base_size = 10)
```


# Discussion

In this study, we have endeavored to illustrate the practical
consequences of visit placement in longitudinal clinical trials.
Both designs use the same number of subjects and the same number
of visits; they differ only in the timing of observations.

The results are broadly consistent, in direction, with the
theoretical findings of @winkensOptimalTimepointsClinical2005,
but not through the mechanism that paper's compound-symmetry
analysis invokes. As derived in Section
\@ref(covariance-structure-and-theoretical-prediction), the
data-generating mechanism used here is the standard
random-intercept, random-slope covariance structure, not a
structure that "combines features of compound symmetry and AR(1)
with measurement error" as an earlier draft of this report
incorrectly stated; it contains no autoregressive component. The
clustered design's precision advantage for the slope contrast
$\beta_3$ follows instead from the classical leverage argument for
regression slopes. The advantage is not an artifact of Monte
Carlo noise: the paired bootstrap relative efficiency (ratio of
empirical variances) is `r sprintf("%.2f", re_boot$point)` (95%
bootstrap interval `r sprintf("%.2f", re_boot$ci[1])` to
`r sprintf("%.2f", re_boot$ci[2])`), consistent with the analytic
relative efficiency of `r sprintf("%.3f", analytic_re)` from
Section \@ref(analytic-relative-efficiency), and the interval
excludes 1.

Several caveats warrant mention. First, we assumed complete data
with no dropout. In practice, attrition is common in
longitudinal trials and may differentially affect designs that
place observations at extreme time-points; adding a dropout
mechanism to the simulation requires a design decision (MCAR/MAR,
rate, whether attrition is tied to visit timing) not made here.
Second, the main simulation's data-generating model used
independent random effects and homoscedastic residuals, but
Sections \@ref(scenario-extension-analytic-only) and
\@ref(scenario-extension-monte-carlo) show, both analytically and
by Monte Carlo simulation, that the clustered design's precision
advantage is not an artifact of that specific structure: it holds
under compound symmetry and under AR(1) plus measurement error as
well, and across four visit schedules spanning a range of boundary
clustering, although the advantage shrinks and became
indistinguishable from Monte Carlo noise in one of 15 non-baseline
cells as the random-slope variance grew large relative to the
residual variance. Heterogeneous (non-constant) residual variances
were not examined and could still alter the relative performance of
the two designs. Third, we considered only a linear treatment
effect; @winkens2007polynomial showed that optimal designs differ
for polynomial effects of higher order.

In conclusion, three points merit emphasis. First, the placement
of observation times in a longitudinal trial is a design
decision with real consequences for statistical efficiency, and
the default of equal spacing is not in general optimal. Second,
the magnitude of the clustering advantage depends on the covariance
structure and on how large the random-slope variance is relative to
the residual variance, but across the structures and schedules
examined here (random-slope, compound symmetry, and AR(1) plus
measurement error; Sections
\@ref(scenario-extension-analytic-only)-\@ref(scenario-extension-monte-carlo))
the qualitative advantage of boundary clustering for the linear
slope contrast is not specific to one structure. Third, the
robustness of these findings to dropout, heterogeneous residual
variances, and nonlinear treatment trajectories remains to be
investigated. The maximin D-optimal framework of
@ouwens2002maximin and the Bayesian approach of @abebe2014bayesian
provide principled starting points for that work.


# References


::: {#refs}
:::
